\forestclaw: Hybrid forest-of-octrees AMR for hyperbolic conservation laws

### 2013-06-07 docs/parco13/paper.tex
Add initial revisions and todo list to PARCO13

@@

\begin{abstract}
%
We present a new hybrid paradigm for parallel adaptive mesh refinement (AMR)
that combines the scalability and lightweight architecture of tree-based AMR
with the computational efficiency of patch-based solvers for hyperbolic
conservation laws.  The key idea is to interpret each leaf of the AMR hierarchy
as one uniform compute patch in $\sR^d$ of dimensions $m^d$, where $m$ is
customarily between 8 and 32.  Thus, computation on each patch can be optimized
for speed, while we inherit the flexibility of adaptive meshes.  In our work we
choose to work with the \pforest AMR library since it allows us to compose the mesh
from multiple mapped octrees and enable the cubed sphere and other
nontrivial multiblock geometries.
%
\end{abstract}

@@

Consequently, it is not clear a priori how to accelerate PDE solvers
that use adaptive mesh refinement.
%, especially when working with
%unstructured meshes \comment{This sounds we are now going to talk
%  about unstructured meshes.}.
% CB: Good point, took it out.
Of course, it was realized early that it helps to aggregate degrees of
freedom (DOF) at the element level, as has been done with high-order
spectral element \cite{TufoFischer99}, low order continuous Galerkin via methods that accumulate many elements simultaneously \cite{knepleyterrel:2013},
%finite volume (element?)
or discontinuous Galerkin \cite{HesthavenWarburton02}
methods.  GPU implementations of the latter have been proposed recently
\cite{KlocknerWarburtonBridgeEtAl09, BursteddeGhattasGurnisEtAl10}.
The finite volume method has typically been implemented using a single degree
of freedom per cell on
% either
structured
\cite{ppm, clawpack}
or unstructured meshes
\cite{openfoam},
where higher order methods have been constructed by widening the stencil, for instance in WENO methods
\cite{Shu:2009bi}.
% \comment{Are you looking for
%   references to parallel AMR codes?  Or more general AMR codes based
%   on the FVM?}  \todo{No, I was looking for unstructured FV
%   references.  Would you have any, and can we add the block-AMR
%   references later?} \comment{But most finite volume methods
%   (structured or unstructured) only have a single degree of freedom
%   per mesh cell (``element''?)  Only recently have people increased
%   the order by extending the stencil (without increase the DOFs per
%   mesh cell).}
% CB: Can you fix some of the above citations?

@@

To facilitate hardware acceleration for parallel dynamic AMR, we
build upon the forest-of-octrees paradigm because of its low overhead
and proven scalability \cite{BursteddeWilcoxGhattas11}.  This approach
identifies each octree leaf with a mesh element.  In our work, we go
beyond the traditional high-order element and define each element to
be a dense computational patch with $m^d$ DOFs.  In fact, this approach
resembles block-structured AMR \cite{be-ol:1984, be-co:1989,
% be-le:1991 (cited below)
ColellaGravesKeenEtAl07,
BerzinsLuitjensMengEtAl10} except that the
patches are not overlapping,
% (except in ghost cell regions),
% CB: we comment on this later in Section 4.
which enables us to capitalize on our
previous experience with scalable FE solvers for PDEs
\cite{BursteddeStadlerAlisicEtAl13}.  The \clawpack software
\cite{LeVeque97} provides a popular implementation of such a patch.
It has been designed to solve hyperbolic conservation laws
% on a uniform compute patch
and successfully used in the context of block-structured AMR
\cite{be-le:1991, amrclaw, Berger:2011du}.

@@

A main challenge is presented by the fact that the patch neighborhood
is only known to \pforest.  This patch connectivity information needs
to be propagated to the numerical code in \forestclaw that implements
the interaction with neighbor patches via the use of a layer of ghost
cells surrounding each patch.  To this end, we define an interface
that allows read-only access to the sequence of blocks,
the list of patches for each, and a lookup of neighbor patches (and
their relative orientation which is nontrivial between blocks).
Suitably informed by \pforest, \forestclaw stores only the patches local to
each MPI rank.  The exchange of ghost data and parallel repartitioning is
provided transparently by \pforest.  Mesh modification directives, such as
adaptive refinement and coarsening, are called from \forestclaw and relayed to
\pforest.

@@

\forestclaw uses iterators over all rank-local leafs, optionally restricted to
a given level.  Random access is possible and used when executing $O(1)$ time
neighbor lookups.  Looping over the patches in the order prescribed by the
forest
% \comment{tree?}
% CB: I think forest is fine here.
and
accessing neighbors only relative to the current patch leads to a high
percentage of cache reuse
\cite{BursteddeBurtscherGhattasEtAl09}.
% since the ghost patches are likely closeby with respect to the $z$-order.

When \forestclaw accesses neighbor patches, they can be on the same or a
different block.  In the latter case, coordinate transformations are carried
out.  The structure of \forestclaw is oblivious to the fact that it only has a
local view of the distributed mesh and data which relieves the developer from
programming to the MPI interface.  Parallel neighbor exchanges are hidden
inside \pforest and called by \forestclaw at well-defined synchronization
points.  There is one such parallel data exchange per time step for a global
value of $\dt$, or one exchange per discretization level per time step if $\dt$
is chosen per-patch depending on its size (this is sometimes called
sub-cycling).  When using sub-cycling, the load balance is attained per level as
it is done for example in Chombo \cite{ColellaGravesKeenEtAl07}
or recent geometric adaptive multigrid schemes
\cite{SundarBirosBursteddeEtAl12}.

The threaded parallelism over the degrees of freedom of a patch can be handled
by \forestclaw alone without the need to involve \pforest.  Additionally many-core implementations, such as \manyclaw \cite{manyclaw}, can be used for the hyperbolic integration on a leaf-patch.  The design of leaf-patches allows for the smart management of data local to accelerator and the host.

>>>>

\newcommand{\alert}[1]{{\color{red}#1}}

@@

\begin{abstract}
We present a new hybrid paradigm for parallel adaptive mesh
refinement (AMR) that combines the scalability and lightweight architecture of
tree-based AMR with the computational efficiency of patch-based solvers for
hyperbolic conservation laws.  The key idea is to interpret each leaf of the AMR
hierarchy as one uniform compute patch in $\sR^d$ with $m^d$ degrees of freedom,
where $m$ is customarily between 8 and 32.  Thus, computation on each patch can
be optimized for speed, while we inherit the flexibility of adaptive meshes.  In
our work we choose to work with the \pforest AMR library since it allows us to
compose the mesh from multiple mapped octrees and enables the cubed sphere and
other nontrivial multiblock geometries.
\end{abstract}

@@

Consequently, it is not clear a priori how to accelerate PDE solvers
that use adaptive mesh refinement.
%, especially when working with
%unstructured meshes \comment{This sounds we are now going to talk
%  about unstructured meshes.}.
% CB: Good point, took it out.
Of course, it was realized early that it helps to aggregate degrees of
freedom (DOF) at the element level, as has been done with high-order
spectral element \cite{TufoFischer99}, low order continuous Galerkin methods that accumulate many elements simultaneously \cite{knepleyterrel:2013},
%finite volume (element?)
or discontinuous Galerkin \cite{HesthavenWarburton02}
methods.  GPU implementations of the latter have been proposed recently
\cite{KlocknerWarburtonBridgeEtAl09, BursteddeGhattasGurnisEtAl10}.
The finite volume method has typically been implemented using a single degree
of freedom per cell on
% either
structured
\cite{ppm, clawpack}
or unstructured meshes
\cite{openfoam};
higher order methods have also been constructed by widening the stencil, for instance in WENO methods
\cite{Shu:2009bi}.
% \comment{Are you looking for
%   references to parallel AMR codes?  Or more general AMR codes based
%   on the FVM?}  \todo{No, I was looking for unstructured FV
%   references.  Would you have any, and can we add the block-AMR
%   references later?} \comment{But most finite volume methods
%   (structured or unstructured) only have a single degree of freedom
%   per mesh cell (``element''?)  Only recently have people increased
%   the order by extending the stencil (without increase the DOFs per
%   mesh cell).}
% CB: Can you fix some of the above citations?

@@

To facilitate hardware acceleration for parallel dynamic AMR, we
build upon the forest-of-octrees paradigm because of its low overhead
and proven scalability \cite{BursteddeWilcoxGhattas11}.  This approach
identifies each octree leaf with a mesh element.  \alert{The presented work goes beyond the traditional high-order element and defines each element to
be a dense computational patch with $m^d$ DOFs.}  In fact, this approach
resembles block-structured AMR \cite{be-ol:1984, be-co:1989,
% be-le:1991 (cited below)
ColellaGravesKeenEtAl07,
BerzinsLuitjensMengEtAl10} except that the
patches are not overlapping,
% (except in ghost cell regions),
% CB: we comment on this later in Section 4.
which enables us to capitalize on our
previous experience with scalable FE solvers for PDEs
\cite{BursteddeStadlerAlisicEtAl13}.  The \clawpack software
\cite{LeVeque97} provides a popular implementation of such a patch.
It has been designed to solve hyperbolic conservation laws
% on a uniform compute patch
and successfully used in the context of block-structured AMR
\cite{be-le:1991, amrclaw, Berger:2011du}.
\comment{Review 1 points out that the statement in red may not be what we are intending to say here.  Mainly the distinction between claiming better convergence then a high-order method (which I think is not what we are saying) and just stating that our elements are patches themselves as opposed to a single DOF.}

@@

A main challenge is presented by the fact that the patch neighborhood
is only known to \pforest.  This patch connectivity information needs
to be propagated to the numerical code in \forestclaw that implements
the interaction with neighbor patches via the use of a layer of ghost
cells surrounding each patch.  \alert{To this end}, we define an interface
that allows read-only access to the sequence of blocks,
the list of patches for each, and a lookup of neighbor patches (and
their relative orientation which is nontrivial between blocks).
Suitably informed by \pforest, \forestclaw stores only the patches local to
each MPI rank.  \alert{The exchange of ghost} data and parallel repartitioning is
provided transparently by \pforest.  Mesh modification directives, such as
adaptive refinement and coarsening, are called from \forestclaw and relayed to
\pforest.

\alert{Review 2 requests more info on how this is done.  Sounds like they are more curious about the refinement overhead to other leaves on the tree and the strategy as to how refinement is decided.}

@@

\forestclaw uses iterators over all rank-local leaves, optionally restricted to
a given level.  Random access is possible and used when executing $O(1)$ time
neighbor lookups.  Looping over the patches in the order prescribed by the
forest
\alert{Review 2 - Explain the preceding paragraph in more detail}
% \comment{tree?}
% CB: I think forest is fine here.
and
accessing neighbors only relative to the current patch leads to a high
percentage of cache reuse
\cite{BursteddeBurtscherGhattasEtAl09}.
% since the ghost patches are likely closeby with respect to the $z$-order.

When \forestclaw accesses neighbor patches, they can be on the same or a
different block.  In the latter case, coordinate transformations are carried
out.  The structure of \forestclaw is oblivious to the fact that it only has a
local view of the distributed mesh and data which relieves the developer from
programming to the MPI interface.  Parallel neighbor exchanges are hidden
inside \pforest and called by \forestclaw at well-defined synchronization
points.  There is one such parallel data exchange per time step for a global
value of $\dt$, the time step length, or one exchange per discretization level per time step if $\dt$
is chosen per-patch depending on its size (this is sometimes called
sub-cycling).  When using sub-cycling, the load balance is attained per level as
it is done for example in Chombo \cite{ColellaGravesKeenEtAl07}
or recent geometric adaptive multigrid schemes
\cite{SundarBirosBursteddeEtAl12}.

The threaded parallelism over the degrees of freedom of a patch can be handled
by \forestclaw alone without the need to involve \pforest.  Additionally many-core implementations, such as \manyclaw \cite{manyclaw}, can be used for the integration of the hyperbolic system on a leaf-patch thereby allowing for hybrid parallelism.  The design of leaf-patches then can enable efficient management of data local to many-core architectures (e.g. the Xeon Phi), and the host.
